---
title: "rogersmodel_vtk"
author: "Brendan Barrett"
date: "4/28/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Alan Rogers' Model

We wish to examine if and how social learning can evolve in temporally varying environments by analyzing a continuous environment version of the [Rogers' Model](https://www.jstor.org/stable/27563823?seq=1). 
We wish to more formally evaluate our verbal model that social learning is adaptive in changing environments, and that social learning evolves because it minimizes the costs of individual learning.

For this we will specify the fitness function for each genotype (in this case individual or social learner).
Then we will make assertions about the life cycle of the individual.
Then we input information about the environment. 

## Environmental Conditions
The environment changes continuously with a probability of $u$. This can be thought of flipping a (likely biased) coin (bernouli choice) where the probability of landing on one side is $u$.  that shows a stable environment, where $0 \leq u \leq 1$. 
If u is equal to 1, the environment changes every generation and is never stable. 
If $u$ = 0 the environment is stable. 
Agents do not know if the environment has changed.


## Life cycle of organism.
We assume a haploid organism with nearly non-overlapping generations. 
Parents give birth to offspring and live with just enough time that they may act as demonstrators to juveniles. 
Then parents die.

## Individual Learners

Individual learners do not use social information or care about environmental state. 
They exist at a frequency of $1-p$.
Their fitness function is:
\begin{equation}
  \label{eq:Wi}
W(I) = w_0 + b - c
\end{equation}

$w_0$ is baseline fitness. It is the fitness of everything else aside from the behavior of interest.

$b$ be the benefit of performing adaptive behavior. In this model we assume that individual learners always come up with adaptive behavior.

$c$ is the cost of individual learning. It can be thought of as a sampling cost.

Individual learners are born, innovate a behavior at a cost, reproduce, and die.

## Social learners
Social learners exist in the population with a frequency of $p$. 
When they are born they randomly copy a member of the previous generation (called oblique transmission), right before the previous generation dies. 
Their fitness function is the time varying recursion equation below:

\begin{equation}
  \label{eq:Ws}
W_t(S) = w_0 + Q_t b
\end{equation}

Note that this fitness expression is time varying (denoted by the little $t$). $Q_t$ is the frequency of adaptive behavior in the population at timestep $t$. 

Let $Q_t$ be the probability of acquiring adaptive behavior by social learning in generation $t$. 
$Q_t$ is a recursion equation. 
It changes over time.
We define $Q_t$ as:
\begin{equation}
	Q_t = (1-u)\big( (1-p) + pQ_{t-1} \big) + u(0) \label{Qt}
\end{equation}

To estimate the frequency of adaptive behavior we have to sum up all the possible combinations of genotypes and environmental interactions that are possible. Lets assume the environment has not changed:

1. Social learners may copy social learners with adaptive behavior: $(1-u) p Q_{t-1}$

2. Social learners may copy individual learners is the environment does not change: $(1-u) (1-p)$

If the environment changes, there is not fitness benefit to social learning, as all previous behavior is non-adaptive regardless if the demonstrator was an individual or social learner. Hence we get $u(0)$.

In the right hand side of the equation we see $Q_{t-1}$. This is the frequency of adaptive behavior in the previous generation. Thus this is a recursive, time-varying model where each timestep $t$ is a generation.

## Solve equilibrium conditions.
We wish to solve for the steady distribution of socially learned behaviors in the population and find the equilibrium value, $\bar Q$. This means that the frequency of social learners will not change. To do so, we may solve the simultaneous system of equations:
\begin{equation}
W_t(S) = W(I)
\end{equation}

\begin{equation}
\hat Q = Q_t = Q_{t-1}
\end{equation}


## The Algebra
For $\hat p = p$ and $\hat Q = Q_{t-1} = Q_t$ there is a unique solution for the equilibrium values (denoted by hats) of the frequency of social learners and adaptive behavior:

First lets solve for the steady staty of adaptive behavior $\bar Q$:

\begin{equation}
\bar Q = (1-u) (1 - p + p \bar Q )
\end{equation}

\begin{equation}
\bar Q = (1-u) (1 - p) + (1-u) p \bar Q )
\end{equation}

\begin{equation}
\bar Q - (1-u) p \bar Q  = (1-u) (1-p) 
\end{equation}

\begin{equation}
\bar Q (1 -p (1-u)) = (1-u) (1-p) 
\end{equation}

\begin{equation}
\bar Q = \frac{(1-u) (1-p)}{1 -p (1-u)}
\end{equation}

With this value, we can estimate the fitness of social leaners at the stationaty value of $\bar Q$. 
This is:

\begin{equation}
\hat W(S) = w_0 + b \bar Q = w_0 + b \frac{(1-u) (1-p)}{(1 -p (1-u))}
\end{equation}


At this stationary value, we can solve for the conditions where individual and social learners coexist at equalibrium:

First lets solve for $\hat Q$ the equilibrium frequency of adaptive behavior.
\begin{equation}
\hat W(S)=W(I)
\end{equation}

\begin{equation}
w_0 + b \hat Q = w_0 + b -c
\end{equation}

\begin{equation}
b \hat Q =  b - c
\end{equation}

\begin{equation}
\hat Q = 1 - c/b
\end{equation}

Next lets solve for  $\hat p$, the equilibrium frequency of social learners:
\begin{equation}
\hat W(S)=W(I)
\end{equation}

\begin{equation}
w_0 + b \frac{(1-u) (1-p)}{1 -p (1-u)} = w_0 + b - c
\end{equation}

\begin{equation}
\frac{b(1-\hat p)(1-u)}{1-\hat p(1-u)} = b - c
\end{equation}

\begin{equation}
(1-\hat p)(1-u) = (b - c)(1-\hat p (1-u))
\end{equation}

\begin{equation}
(b-b \hat p)(1-u) = b - b \hat p (1-u) -c +c \hat p (1-u) 
\end{equation}

\begin{equation}
(b-b \hat p)(1-u) = b - b \hat p (1-u) -c + c \hat p (1-u) 
\end{equation}

Now I get lazy and drop the hat.

\begin{equation}
b - ub -bp +ubp = b - bp + bpu - c + cp -cpu
\end{equation}

\begin{equation}
-ub = -c + pc - pcu
\end{equation}

\begin{equation}
c-ub =  p(c - cu)
\end{equation}

\begin{equation}
c-ub =  p(c(1 - u))
\end{equation}

\begin{equation}
 p= \frac{c-ub}{c(1 - u)}
\end{equation}

\begin{equation}
	\hat p = \frac{1 - u b/c}{1-u} 
\end{equation}

## Evaluating Invasion Dynamics

### Invasion of individual learners
Examining the invasion dynamics by setting $p \approx 0$ we find that social learners may invade a population of individual learners when:
\begin{equation}
W(I) > \hat W (S)
\end{equation}

Simplifying shows this is:
\begin{equation}
u<c/b
\end{equation}


Setting  $p \approx 1$, individual learners may invade a population of social learners when $W(I) < \hat W (S)$ or:

\begin{align*}
b>c
\end{align*}

### Comparing fitness functions of 2 strategies
```{r fitness graph , echo=TRUE , warning=FALSE}
b <- 3
c <- 1
w0 <- 1
u <- 0.15
p_seq <- seq(from=0 , to=1 , by=0.01)
wI <- wS <- rep (NA , length(p_seq) )

for( i in 1:length(p_seq) ){
  Qbar <- (1-u)*(1-p_seq[i]) / ( 1 - p_seq[i]*(1-u))
  wI[i] <- w0 + b - c
  wS[i] <- w0 + b*Qbar
}
plot(wS ~ p_seq , pch=19 , col="orange" , ylab="fitness" , xlab="p: frequency of social learners")
points(wI ~ p_seq ,  pch=19 , col="slateblue" )
```  


## Simulating Rogers Model in R

We can translate this model into a simulation. First what we will do is specify the initial conditions
```{r rogersinit conds, echo=TRUE , warning=FALSE}
#set initial parameter conditions
w0 <- 1 #baseline fitness
b <- 3 #benefit of adaptive behavior
c <- 1 #cost of social learning
U <- 0.05 #probability of environmental change
#s <- 0.9 #success of innovation
timesteps <- 500
#wI <- w0 + b - c
```

```{r rogers loop code, echo=TRUE , warning=FALSE}
#containers for values
q <- p <- u <-  rep(0,timesteps + 1 )
wS <- wI  <- wPOP <-  rep(0,timesteps )

#calculate phat from empirical model
phat <- (1-U*b/c)/(1-U) #equilibrium values for frequency of social learners

#set starting conditions
q[1] <- 0 
p[1] <- phat 
p[1] <- 0.001 
u[1] <- 0

#begin simulation
for (t in 1:timesteps){
  wI[t] <- w0 + b - c #fitness individual learners
  wS[t] <- w0 + b*q[t] #fitness of social learners
  wPOP[t] <- wI[t]*(1-p[t]) + wS[t]*p[t] #population mean fitness
  q[t+1] <- (1-u[t])*( (1-p[t]) + p[t]*q[t]) + u[t]*0 #frequecncy of adaptive behavior
  p[t+1] <- p[t]*wS[t]/(p[t]*wS[t] + (1-p[t])*wI) #frequecny of social learners
  u[t+1] <- rbinom(n=1,prob=U,size=1) #sample environment changing in next gen, if 0 all SL behav maladaptive
}
```

Next we can plot the evolutionary dynamics of our simulation:

```{r rogers plot code, echo=TRUE , warning=FALSE}

par(mfrow = c(3, 1) )
par(cex = 0.4)
par(oma = c(4,2,1,1) )
par(mar = c(2,4,0,0.1) )

plot( p ~ c(1:(timesteps + 1) ) , ylim=c(-0.02,1) , pch=19 , col="white" , ylab="frequency of social learners (p)" , cex.lab=1.5)
lines( (1:(timesteps +1)) , p , col="red", lty=1, lwd=2)
points(  seq(0:timesteps) , (u-1) , pch=17)

plot( q ~ c(1:(timesteps + 1) ) , ylim=c(-0.02,1) , pch=19 , col="white" , ylab="frequency of adaptive behavior" , cex.lab=1.5)
lines( (1:(timesteps +1)) , q , col="blue", lty=1, lwd=2)
points(  seq(0:timesteps) , (u-1) , pch=17)

plot( wS ~ c(1:(timesteps) ) , ylim=c(0.9,b+w0) , pch=19 , col="white" , ylab="phenotypic fitness" , cex.lab=1.5)
lines( c(1:timesteps) , wS , col="orange", lty=1, lwd=2)
lines( c(1:timesteps ) , wI , col="green", lty=1, lwd=2)
lines( c(1:timesteps ) , wPOP , col="black", lty=1, lwd=2)
legend("topleft", legend=c("SL fitness (wS)", "IL fitness (wI)", "mean population fitness"),
       col=c("orange", "green" , "black"), lty=1, cex=1.2 , lw=2 , bty="n")
points(  seq(0:timesteps) , (u) , pch=17)

##Have class show that phat in simulation is the same as phat in equation
##Show invasion conditions of SL via simulation (when they can and can't invade)
##Show invasion conditions of IL via simulation (where they can and can't invade)
```
# Exercises
### Exercise 1
Convert this code into a function in $R$, where you can specify parameter values of $b$, $c$, and $u$, and initial frequency of $p$ and it spits out the above graph.

1. Show a simulation where individal learners can invade.

2. Show a simulation where social learners can invade.

3. Show a simualtion where social learners cannot invade.

### Exercise 2
We assume that innovation is always succesful for individual learners. This is not likely the case. Lets make innovators fail sometimes by introducing a new parameter, $s$, the success of innovation. This takes on a value between 0 and 1.

We can add this by changing the fitness functions of individual learners to: 

\begin{equation}
W(I) = w_0 + s b - c
\end{equation}

which then changes the recursion for $Q_t$:
\begin{equation}
	Q_t = (1-u)\big( (1-p)s + pQ_{t-1} )
\end{equation}

  1. Show how innovation errors affect the steady state frequency of adaptive behavior and frequency of social learners. You can do this via simulation, alegraically, or both.

### Exercise 3
One of the phenomenon we can see in this model is that genetic fitness can become decoupled from adaptive behavior. 
1. Run a simulation(s) that shows this is the case. What are the conditions where this can happen?